Concurrent Programming with CXL

### 2020-06-20 book/paper.tex
Fixed synchS

\begin{itemize}
\item Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter.
\item Chapter~\ref{ch:cxlmachine} presents the
CXL virtual machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces hardware-supported locks
for implemented critical sections.  Chapter~\ref{ch:synch} presents a CXL module
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} describes \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item Chapter~\ref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches processes may not be able to get access to a
resource they need.
\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present
\emph{monitors} and \emph{condition variables},
another approach to process synchronication.
\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
avoid them.
\item Chapter~\ref{ch:actor} presents the \emph{actor model} as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

\chapter{Locks and the Synch Module}
\label{ch:synch}
\index{lock}
\index{synch module}

@@

Locks are probably the most popular and basic form of synchronization
in concurrent programs.  For this reason, CXL has a module called
\texttt{synch} that includes support for locks.
Figure~\ref{fig:spinlocks} shows how they are implemented, and
Figure~\ref{fig:incfixed} gives an example of how they may be used,
in this case to fix the program of Figure~\ref{fig:inc}.
Notice that the module completely hides the implementation of the
lock.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.

@@

In most operating systems, processes are virtual and can be suspended
until some condition changes.  For example, a process can be suspended
until the lock is available to it.
In CXL, a process can suspend itself and save its context in a
list.  Recall that the context of a process consists of its name tag,
its program counter, and the contents of its register and stack.
A context is a regular CXL value.
The syntax of the expression is as follows:

@@

There is a second version of the \texttt{synch} module that uses suspension
instead of spinlocks.
Figure~\ref{fig:suspension} shows the same \texttt{Lock} interface implemented
using suspension.
A \texttt{Lock} maintains both a boolean indicating whether the lock is taken,
and a list of contexts of processes that want to acquire the lock.
\texttt{lock} sets the lock if available, or suspends itself if not.
Note that \texttt{stop} is called within an \texttt{atomic} statement---this is
the only exception to an atomic statement running to completion.  While the
process is running no other processes can run, but when it suspends itself
it allows other processes to run.

\texttt{unlock} checks to see if any processes are waiting to get the lock.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module to resume the first process that got
suspend and to remove its context from the list.
Selecting the first process is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents \emph{starvation}:
\index{starvation}
every process
gets a chance to obtain the lock (assuming every process eventually releases
it).  Chapter~\ref{ch:starvation} will talk more about starvation.

@@

runs the file \texttt{x.cxl} using the Suspension version of the \texttt{synch} module.
You will find that using the \texttt{synchS} module often leads to searching a
significantly larger search space than using the \texttt{synch} module.
Part of the reason is that the \texttt{synchS} module keeps track of the order
in which processes wait for a lock.

@@

\chapter{List of Modules}

>>>>

\begin{itemize}
\item Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter.
\item Chapter~\ref{ch:cxlmachine} presents the
CXL virtual machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces ``hardware'' locks
for implemented critical sections.
\item Chapter~\ref{ch:synch} presents CXL modules
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} describes \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item Chapter~\ref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches processes may not be able to get access to a
resource they need.
\item Chapters~\ref{ch:monitors} and~\ref{ch:mesa} present
\emph{monitors} and \emph{condition variables},
another approach to process synchronication.
\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
avoid them.
\item Chapter~\ref{ch:actor} presents the \emph{actor model} as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

\chapter{Locks and Blocking}
\label{ch:synch}
\index{lock}
\index{synch module}

@@

Locks are probably the most prevalent and basic form of synchronization
in concurrent programs.  For this reason, CXL has a module called
\texttt{synch} that includes support for locks.
Figure~\ref{fig:spinlocks} shows how they are implemented, and
Figure~\ref{fig:incfixed} gives an example of how they may be used,
in this case to fix the program of Figure~\ref{fig:inc}.
Notice that the module completely hides the implementation of the
lock.
The \texttt{synch} module includes a variety of other useful
synchronization primitives, which will be discussed in later
chapters.

@@

In most operating systems, processes are virtual and can be suspended
until some condition changes.  For example, a process can be suspended
until the lock is available to it.
In CXL, a process can suspend itself and save its context (state) in a
list.  Recall that the context of a process consists of its name tag,
its program counter, and the contents of its register and stack.
A context is a regular CXL value.
The syntax of the expression is as follows:

@@

There is a second version of the \texttt{synch} module that uses suspension
instead of spinlocks.
Figure~\ref{fig:suspension} shows the same \texttt{Lock} interface implemented
using suspension.
The interface is implemented as follows:
\begin{itemize}
\item A \texttt{Lock} maintains both a boolean indicating whether the
lock is taken and a list of contexts of processes that want to acquire the lock.
\item
\texttt{lock} sets the lock if available and suspends itself if not.
Note that \texttt{stop} is called within an \texttt{atomic} statement---this is
the only exception to an atomic statement running to completion.  While the
process is running no other processes can run, but when it suspends itself
it allows other processes to run.
\item
\texttt{unlock} checks to see if any processes are waiting to get the lock.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module (see Appendix~\ref{ap:module})
to resume the first process that got
suspended and to remove its context from the list.
\end{itemize}
Selecting the first process is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents \emph{starvation}:
\index{starvation}
every process
gets a chance to obtain the lock (assuming every process eventually releases
it).  Chapter~\ref{ch:starvation} will talk more about starvation and how
to prevent it.

@@

runs the file \texttt{x.cxl} using the suspension version of the \texttt{synch} module.
You will find that using the \texttt{synchS} module often leads to searching a
significantly larger search space than using the \texttt{synch} module.
Part of the reason is that the \texttt{synchS} module keeps track of the order
in which processes wait for a lock.

@@

\chapter{List of Modules}
\label{ap:module}
